Parametrization of Neural Nets with Connected Abelian Lie Groups as Data Manifold

### 2020-05-06 llncs2e/paper77.tex
net

\usepackage[hidelinks]{hyperref}
\hypersetup{
    pdftitle={Parametrization of Neural Networks with Connected Abelian Lie Groups as Data Manifold},    % title
    pdfauthor={Luciano Melodia},     % author
    pdfsubject={Neural Networks on Connected Abelian Lie Groups},   % subject of the document
    pdfcreator={Luciano Melodia},   % creator of the document
    pdfproducer={Luciano Melodia},  % producer of the document
    pdfkeywords={Embedding Dimension} {Parametrization} {Persistent Homology} {Neural Nets} {Manifold Learning}, % list of keywords
    pdfnewwindow=true,
    colorlinks,
	citecolor=black,
	filecolor=black,
	linkcolor=black,
	urlcolor=blue
}

@@

\tikzset{->-/.style={decoration={markings, mark=at position #1 with {\arrow[line width=1pt]{>}}},postaction={decorate}}}
\begin{document}
%
\title{Parametrization of Neural Networks with Connected Abelian Lie Groups as Data Manifold}
%
\titlerunning{Neural Networks on Connected Abelian Lie Groups}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Luciano Melodia\orcidID{0000-0002-7584-7287} \\
\and Richard Lenz\orcidID{0000-0003-1551-4824}}
%
\authorrunning{L. Melodia et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Chair of Computer Science 6\\
Friedrich-Alexander University Erlangen-Nürnberg\\
91058 Erlangen, Germany \\
\email{\{luciano.melodia,richard.lenz\}@fau.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Neural nets have been used in an substantial number of scientific disciplines. Despite this, their parameterization is largely unexplored. Dense nets are the coordinate transformations of a manifold from which the data is sampled. After a layer processes the data, the representation of the original manifold may change. This is crucial for the preservation of its topological structure and should therefore be parameterized correctly. We discuss a method to determine the smallest topology preserving layer considering the data domain as abelian connected Lie group and observe that it is decomposable into $\mathbb{R}^p \times \mathbb {T}^q$. Persistent homology allows us to count its $k$-th homology groups. Using Künneth's theorem, we count the $k$-th Betti numbers. Since we know the embedding dimension of $\mathbb{R}^p$ and $\mathcal{S}^1$, we parameterize the bottleneck layer with the smallest possible matrix group, which can represent a manifold with those homology groups. Resnets guarantee smaller embeddings due to the dimension of their state space representation.

@@

\subsection{Residual Neural Networks}
\label{resnet}
A \textit{resnet} skips some of the layers by adding outputs to a deeper layers. It has one \textit{skip connection} and is a $\mathcal{C}^1$-continuous dynamic transformation \citemain{he2016deep,HauserGJR19}:
\begin{align}
	\label{residualnet}
	\textbf{x}^{(l+1)} = \textbf{x}^{(l)} + \sigma^{(l)}\left(\textbf{x}^{(l)}\right) \Delta l.
\end{align}
The term $\Delta l$ in Eq. \ref{residualnet} is inserted as a perturbation term. In the work of Hauser et al. \citemain{HauserR17}, neural nets were classified as $\mathcal{C}^k$-architecture, depending on how often a \textit{finite difference operator} is applied to the map $\textbf{x}: M \times I \rightarrow \mathbb{R}^d$. The forward and backward difference operators are defined as $\delta^{+}: \textbf{x}^{(l)} \mapsto \textbf{x}^{(l+1)} - \textbf{x}^{(l)}$ and $\delta^{-}: \textbf{x}^{(l)} \mapsto \textbf{x}^{(l)} - \textbf{x}^{(l-1)}$. In this theory, we can write the difference operators for a general $n \in \mathbb{N}$ with explicit specification of the perturbation $\Delta l$ as
\begin{align}
	\label{generalresidual}
	\delta^+\left( \delta^- \right)^{n-1} \textbf{x}^{(l)} = \sum_{l'=0}^{k} \left[ (-1)^{l'} {k \choose l'} \textbf{x}^{(l+1-l')} \right] = \sigma^{(l)} \left(\textbf{x}^{(l)}\right) \Delta l^k.
\end{align}
We use the same notation as in \citemain{HauserGJR19}, where $(\delta^-)^{k-1} := \delta^- \delta^- \cdots \delta^-$ denotes the application of the difference operator $(k-1)$-times and the binomial coefficient ${k \choose l'}$, with $l'$ indicating the index for summation over the layers. Using one forward difference operator and the remaining $(k-1)$ backward difference operators, the next layer $\textbf{x}^{(l+1)}$ is a function of the $k$ previous layers $\textbf{x}^{(l)}, \textbf{x}^{(l-1)}, \cdots, \textbf{x}^{(l-k+1)}$. To calculate the \textit{state space dimension} of equations of degree $k$ as in Eq. \ref{generalresidual}, we define the state as the difference operators of different transformation order of an input vector $\textbf{x}$ in the layer $l$ according to \citemain{HauserGJR19}:
\begin{align}
	\textbf{q}_n^{(l)} := \left( \delta^- \right)^{n-1} \textbf{x}^{(l)} \quad \text{for all} \; n = 1,2,\cdots,k.
\end{align}

@@

\subsection{Results}
Tab. \ref{stats} shows the representatives of homology classes from their respective groups for cifar10 and cifar100. Further, it shows the estimated minimal embedding dimension for a manifold with these homology groups for $\mathcal{C}^1$ to $\mathcal{C}^4$-invertible resnets. The representatives for the $k$-th homology group $\sharp_k^{H}\left(\bar{\mathfrak{L}}\left(\mathfrak{P}(K(X))\right)\right)$ of the respective dataset are counted. Then, we seek for an embedding dimension to represent a manifold with the number of spheres $q$ with these homology groups. A Monte Carlo approach is used to find integer solutions of Eq. \ref{countingequation}. All models below our dimensional treshold show explosions of the gradient, see Fig. \ref{costfunctions}. The dimension of the respective loss function can be identified by the transition \gradient{}, from the lowest tested dimension to the highest, see Fig. \ref{costfunctions}. The models above the treshold remain stable. Higher parameterized tested models are stable. A single net still shows an explosive gradient in Fig. \ref{costfunctions} (b). This is due to the fact that the neural network layers themself are not connected nor abelian and are not explicitly restricted to such a structure. The net showing the deflection of gradient is close to our threshold. We specify the parameterization for an embedding as well as for neural nets of the smoothness degree $1$--$4$.
\input{figure4.tex}

\section{Conclusions}
\label{sec:conclusion}
Based on the theory of Lie groups and persistent homology a method for neural networks was developed to parameterize them with good precision due to the assumption that the data lies on or near by some connected commutative Lie group. This forms an approximate solution for a special case of learning problems, allowing to restrict the domain to a commutative group beeing a connected smooth manifold. The initial problem could be simplified by decomposing the Lie group into a product space $\mathbb{R}^p \times \mathbb{T}^q$. The homology groups of the $q$-torus and the Euclidean space are well known. The application of fully invertible neural nets to resnets demonstrates our results on a different Lie group from Euclidean space. By applying Künneth's theorem the homology groups of the topological factor spaces could be connected with the ones of their product space. Using mean confidence persistence landscapes the homology groups were estimated. A treshold was computed, see Sect. \ref{sec:numericalexperiments}, which is determined by the state space dimension of the resnets according to a representation up to homology equivalence. The mean persistence landscapes are particularly useful for short-lived isolated collections of data points, which in particular generate elements from homology classes at the beginning of the filtration and become visible at the lower left corner of the persistence landscapes, see Fig. \ref{homgroups} (a). The theoretical work was tested with numerical experiments on toy data. We do not limit the processing of the neural net within the layers to an abelian Lie group. Instead, we assume that the net learns some map to a structure close to such an abelian connected Lie group with the estimated homology groups, on which the data could lie. This seems to hold, according to our experiments. Still, this method does only approximate the true data manifold in the case that the data lies already on a product space $\mathbb{R}^p \times \mathbb{T}^q$ or is mapped by the neural network to such a manifold.

>>>>

\usepackage[hidelinks]{hyperref}
\hypersetup{
    pdftitle={Parametrization of Neural Nets with Connected Abelian Lie Groups as Data Manifold},    % title
    pdfauthor={Luciano Melodia},     % author
    pdfsubject={Neural Nets on Connected Abelian Lie Groups},   % subject of the document
    pdfcreator={Luciano Melodia},   % creator of the document
    pdfproducer={Luciano Melodia},  % producer of the document
    pdfkeywords={Embedding Dimension} {Parametrization} {Persistent Homology} {Neural Nets} {Manifold Learning}, % list of keywords
    pdfnewwindow=true,
    colorlinks,
	citecolor=black,
	filecolor=black,
	linkcolor=black,
	urlcolor=blue
}

@@

\tikzset{->-/.style={decoration={markings, mark=at position #1 with {\arrow[line width=1pt]{>}}},postaction={decorate}}}
\begin{document}
%
\title{Parametrization of Neural Nets with Connected Abelian Lie Groups as Data Manifold}
%
\titlerunning{Neural Nets on Connected Abelian Lie Groups}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Luciano Melodia\orcidID{0000-0002-7584-7287} \\
\and Richard Lenz\orcidID{0000-0003-1551-4824}}
%
\authorrunning{L. Melodia et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Chair of Computer Science 6\\
Friedrich-Alexander University Erlangen-Nürnberg\\
91058 Erlangen, Germany \\
\email{\{luciano.melodia,richard.lenz\}@fau.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Neural nets have been used in an substantial number of scientific disciplines. Despite this, their parameterization is largely unexplored. Dense nets are the coordinate transformations of a manifold from which the data is sampled. After a layer processes the data, the representation of the original manifold may change. This is crucial for the preservation of its topological structure and should therefore be parameterized correctly. We discuss a method to determine the smallest topology preserving layer considering the data domain as abelian connected Lie group and observe that it is decomposable into $\mathbb{R}^p \times \mathbb {T}^q$. Persistent homology allows us to count its $k$-th homology groups. Using Künneth's theorem, we count the $k$-th Betti numbers. Since we know the embedding dimension of $\mathbb{R}^p$ and $\mathcal{S}^1$, we parameterize the bottleneck layer with the smallest possible matrix group, which can represent a manifold with those homology groups. Resnets guarantee smaller embeddings due to the dimension of their state space representation.

@@

\subsection{Residual Neural Nets}
\label{resnet}
A \textit{resnet} skips some of the layers by adding outputs to a deeper layers. It has one \textit{skip connection} and is a $\mathcal{C}^1$-continuous dynamic transformation \citemain{he2016deep,HauserGJR19}:
\begin{align}
	\label{residualnet}
	\textbf{x}^{(l+1)} = \textbf{x}^{(l)} + \sigma^{(l)}\left(\textbf{x}^{(l)}\right) \Delta l.
\end{align}
The term $\Delta l$ in Eq. \ref{residualnet} is inserted as a perturbation term. In the work of Hauser et al. \citemain{HauserR17}, neural nets were classified as $\mathcal{C}^k$-architecture, depending on how often a \textit{finite difference operator} is applied to the map $\textbf{x}: M \times I \rightarrow \mathbb{R}^d$. The forward and backward difference operators are defined as $\delta^{+}: \textbf{x}^{(l)} \mapsto \textbf{x}^{(l+1)} - \textbf{x}^{(l)}$ and $\delta^{-}: \textbf{x}^{(l)} \mapsto \textbf{x}^{(l)} - \textbf{x}^{(l-1)}$. In this theory, we can write the difference operators for a general $n \in \mathbb{N}$ with explicit specification of the perturbation $\Delta l$ as
\begin{align}
	\label{generalresidual}
	\delta^+\left( \delta^- \right)^{n-1} \textbf{x}^{(l)} = \sum_{l'=0}^{k} \left[ (-1)^{l'} {k \choose l'} \textbf{x}^{(l+1-l')} \right] = \sigma^{(l)} \left(\textbf{x}^{(l)}\right) \Delta l^k.
\end{align}
We use the same notation as in \citemain{HauserGJR19}, where $(\delta^-)^{k-1} := \delta^- \delta^- \cdots \delta^-$ denotes the application of the difference operator $(k-1)$-times and the binomial coefficient ${k \choose l'}$, with $l'$ indicating the index for summation over the layers. Using one forward difference operator and the remaining $(k-1)$ backward difference operators, the next layer $\textbf{x}^{(l+1)}$ is a function of the $k$ previous layers $\textbf{x}^{(l)}, \textbf{x}^{(l-1)}, \cdots, \textbf{x}^{(l-k+1)}$. To calculate the \textit{state space dimension} of equations of degree $k$ as in Eq. \ref{generalresidual}, we define the state as the difference operators of different transformation order of an input vector $\textbf{x}$ in the layer $l$ according to \citemain{HauserGJR19}:
\begin{align}
	\textbf{q}_n^{(l)} := \left( \delta^- \right)^{n-1} \textbf{x}^{(l)} \quad \text{for all} \; n = 1,2,\cdots,k.
\end{align}

@@

\subsection{Results}
Tab. \ref{stats} shows the representatives of homology classes from their respective groups for cifar10 and cifar100. Further, it shows the estimated minimal embedding dimension for a manifold with these homology groups for $\mathcal{C}^1$ to $\mathcal{C}^4$-invertible resnets. The representatives for the $k$-th homology group $\sharp_k^{H}\left(\bar{\mathfrak{L}}\left(\mathfrak{P}(K(X))\right)\right)$ of the respective dataset are counted. Then, we seek for an embedding dimension to represent a manifold with the number of spheres $q$ with these homology groups. A Monte Carlo approach is used to find integer solutions of Eq. \ref{countingequation}. All models below our dimensional treshold show explosions of the gradient, see Fig. \ref{costfunctions}. The dimension of the respective loss function can be identified by the transition \gradient{}, from the lowest tested dimension to the highest, see Fig. \ref{costfunctions}. The models above the treshold remain stable. Higher parameterized tested models are stable. A single net still shows an explosive gradient in Fig. \ref{costfunctions} (b). This is due to the fact that the neural layers themself are not connected nor abelian and are not explicitly restricted to such a structure. The net showing the deflection of gradient is close to our threshold. We specify the parameterization for an embedding as well as for neural nets of the smoothness degree $1$--$4$.
\input{figure4.tex}

\section{Conclusions}
\label{sec:conclusion}
Based on the theory of Lie groups and persistent homology a method for neural nets was developed to parameterize them with good precision due to the assumption that the data lies on or near by some connected commutative Lie group. This forms an approximate solution for a special case of learning problems, allowing to restrict the domain to a commutative group beeing a connected smooth manifold. The initial problem could be simplified by decomposing the Lie group into a product space $\mathbb{R}^p \times \mathbb{T}^q$. The homology groups of the $q$-torus and the Euclidean space are well known. The application of fully invertible neural nets to resnets demonstrates our results on a different Lie group from Euclidean space. By applying Künneth's theorem the homology groups of the topological factor spaces could be connected with the ones of their product space. Using mean confidence persistence landscapes the homology groups were estimated. A treshold was computed, see Sect. \ref{sec:numericalexperiments}, which is determined by the state space dimension of the resnets according to a representation up to homology equivalence. The mean persistence landscapes are particularly useful for short-lived isolated collections of data points, which in particular generate elements from homology classes at the beginning of the filtration and become visible at the lower left corner of the persistence landscapes, see Fig. \ref{homgroups} (a). The theoretical work was tested with numerical experiments on toy data. We do not limit the processing of the neural net within the layers to an abelian Lie group. Instead, we assume that the net learns some map to a structure close to such an abelian connected Lie group with the estimated homology groups, on which the data could lie. This seems to hold, according to our experiments. Still, this method does only approximate the true data manifold in the case that the data lies already on a product space $\mathbb{R}^p \times \mathbb{T}^q$ or is mapped by the neural net to such a manifold.
